The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2019-06-07 Fsing/FThresholdsPaper/FThresholds.tex
Updates to paper, and to fpt documentation

The \emph{FrobeniusThresholds} package contains methods to
 test candidate values for an $F$-pure threshold, and even $F$-jumping numbers, even in some singular rings.

Consider a  $\mathbb{Q}$-Gorenstein ring $R$ of characteristic $p>0$, whose index not divisible by $p$.
Given a parameter $t \in \mathbb{Q}$ and an element $f$ of $R$, the command {\tt isFPT(t,f)} outputs {\tt true} if $t$ is the $F$-pure threshold of $f$, and {\tt false} otherwise.
Moreover, the function {\tt compareFPT} provides further information about the relationship between $t$ and the $F$-pure threshold of $f$; namely,
{\tt compareFPT(t, f)} returns {\tt-1} , {\tt 0}, {\tt 1} when $t$ is less than, equal to, or greater than, respectively, the $F$-pure threshold of $f$.  Setting the option {\tt IsLocal} to {\tt true} tells the function to consider the $F$-pure threshold at the origin.  The following example illustrates this functionality.

@@

The {\tt fpt} function has the option {\tt UseSpecialAlgorithms}.  When
set to {\tt true} (the default value), {\tt fpt} first checks whether $f$ is a monomial, a diagonal polynomial, a binomial, a form in two variables, or a product of factors in simple normal crossing, in that order.

If $f = x_1^{a_n}\cdots x_n^{a_n}$, then its $F$-pure threshold is simply $\min \{ 1/a_1, \ldots, 1/a_n \}$.  When $f$ is a diagonal polynomial, a binomial, or a form in two variables, 
algorithms of Hern\'andez \cite{HernandezFInvariantsOfDiagonalHyp, HernandezFPureThresholdOfBinomial}, or Hern\'andez and Teixeira \cite{HernandezTeixeiraFThresholdFunctions},  are executed to compute the $F$-pure threshold.  We note that algorithms for computing $F$-pure thresholds of binomials in certain special characteristics appear in \cite{ShibutaTakagiLCThresholds}.

@@

\subsection*{Searching for the $F$-pure threshold when no special algorithm applies}

@@

i14 : fpt(f, Attempts => 1, DepthOfSearch => 3) -- the right-hand

      endpoint (nu+1)/p^e is the F-pure threshold

@@

i17 : fpt(f, Attempts => 2, DepthOfSearch => 3) -- the left-hand endpoint

      nu/(p^e-1) is the F-pure threshold

@@

If neither endpoint is the $F$-pure threshold and {\tt Attempts} is set to more than {\tt 2}, then  additional checks are performed at certain numbers within the interval, chosen according to the criteria determined by the value of the option {\tt GuessStrategy}.  We refer the reader for the documentation of this strategy for more details.

@@

i20 : fpt(f, DepthOfSearch => 3, Attempts => 3) -- an additional check

      sharpens the estimate

        3   1
o20 = {--, --}
       62  18

@@

i21 : fpt(f, DepthOfSearch => 3, Attempts => 4) -- and one more finds

      the answer

@@

If {\tt guessFPT} is unsuccessful and {\tt UseFSignature} is set to {\tt true}, then the {\tt fpt} function proceeds to use the convexity of the $F$-signature function and a secant line argument to attempt to narrow down the interval bounding the $F$-pure threshold.

@@

i22 : f = x^5*y^6*(x + y)^9*(x^2 + y^3)^4;

i23 : fpt(f, DepthOfSearch => 3)

        2   1
o23 = {--, --}
       31  14

o23 : List

i24 : fpt(f, DepthOfSearch => 3, UseFSignature => true)

        181   1
o24 = {----, --}
       2750  14

o24 : List

i25 : numeric o23

o25 = {.0645161, .0714286}

o25 : List

i26 : numeric o24 -- UseFSignature sharpened the estimate a bit

o26 = {.0658182, .0714286}

o26 : List
\end{MyVerbatim}
}
\medspace

When {\tt FRegularityCheck} is set to {\tt true} and no exact answer has been found, a final check is run to verify whether the final lower bound for the $F$-pure threshold is the exact answer (if it has not already been eliminated as a possibility).

i27 : f = (x + y)^4*(x^2 + y^3)^6;

i28 : fpt( f, Attempts => 2, DepthOfSearch => 3 )

        3   13
o28 = {--, ---}
       31  125

o28 : List

i29 : fpt(f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true)

      -- UseFSignature improves the answer a bit

        1   13
o29 = {--, ---}
       10  125

o29 : List

i30 : fpt(f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true,

      FRegularityCheck => true) -- FRegularityCheck finds the answer

       1
o30 = --
      10

o30 : QQ
\end{MyVerbatim}
}
\medspace

The computations performed when {\tt UseFSignature} and {\tt FRegularityCheck} are set to {\tt true} are often slow, and often fail to improve the estimate, and for this reason, these options should be used sparingly.
            It is often more effective to increase the values of {\tt Attempts} or {\tt DepthOfSearch} instead.

i31 : f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;

i32 : timing numeric fpt(f, DepthOfSearch => 3, UseFSignature => true,

      FRegularityCheck => true)

o32 = {.0733061, .0769231}
      -- 2.68899 seconds

o32 : Time

i33 : timing numeric fpt(f, Attempts => 5, DepthOfSearch => 3) -- a better

      answer in less time

o33 = {.075, .0769231}
      -- .893389 seconds

o33 : Time

i34 : timing fpt(f, DepthOfSearch => 4) -- the exact answer in even less
      time

       48
o34 = ---
      625
      -- .361882 seconds

o34 : Time
\end{MyVerbatim}
}
\medspace

@@

i2 : f = (y - (x-1)^2)*y^2; --SNC at the origin, but not globally

@@

i4 : isSimpleNormalCrossing(f, IsLocal=>false)

i5 : g = (y-1)^2+(x-7)^2 --doesn't even pass through the origin

       2    2
o5 = x  + y  - 14x - 2y + 50

o5 : R

i6 : isSimpleNormalCrossing(g)

o6 = true

i7 : isSimpleNormalCrossing(g, IsLocal=>false)

o7 = false

i8 : h = x*y^2*(x+1)^3*(y-1)^4; --SNC everywhere

i9 : isSimpleNormalCrossing(h)

o9 = true

i10 : isSimpleNormalCrossing(h, IsLocal=>false)

o10 = true
\end{MyVerbatim}
}
\medspace

>>>>

The \emph{FrobeniusThresholds} package contains methods to
 test candidate values for an $F$-pure threshold, and even $F$-jumping numbers, even in some singular rings.
Consider a  $\mathbb{Q}$-Gorenstein ring $R$ of characteristic $p>0$, whose index not divisible by $p$.
Given a rational number $t$ and an element $f$ of $R$, the command {\tt isFPT(t,f)} outputs {\tt true} if $t$ is the $F$-pure threshold of $f$, and {\tt false} otherwise.
Moreover, the function {\tt compareFPT} provides further information about the relationship between $t$ and the $F$-pure threshold of $f$; namely,
{\tt compareFPT(t, f)} returns {\tt-1} when $t < \fpt(f)$, {\tt 0} when $t = \fpt(f)$ and {\tt 1} when $t > \fpt(f)$.
Setting the option {\tt IsLocal} to {\tt true} tells the function to consider the $F$-pure threshold at the origin.
The following example illustrates this functionality.

@@

The {\tt fpt} function has the option {\tt UseSpecialAlgorithms}.  When
set to {\tt true} (the default value), {\tt fpt} first checks whether $f$ is a monomial, a diagonal polynomial, a binomial, a form in two variables, or a product of factors in simple normal crossing, in that order. 
When $f$ is a monomial, say, $x_1^{a_n}\cdots x_n^{a_n}$, then its $F$-pure threshold is quickly computed as the minimum among the $1/a_i$.
When $f$ is a a diagonal polynomial, a binomial, a form in two variables, 
algorithms of Hern\'andez \cite{HernandezFInvariantsOfDiagonalHyp, HernandezFPureThresholdOfBinomial}, or Hern\'andez and Teixeira \cite{HernandezTeixeiraFThresholdFunctions},  are executed to compute the $F$-pure threshold (cf. \cite{ShibutaTakagiLCThresholds}).  
Consider examples of each of these cases below.

@@

\subsection*{How {\tt fpt} proceeds when no special algorithm applies}

@@

i14 : fpt(f, Attempts => 1, DepthOfSearch => 3) -- the right-hand
      endpoint (nu+1)/p^e is the F-pure threshold

@@

i17 : fpt(f, Attempts => 2, DepthOfSearch => 3) -- the left-hand endpoint
      nu/(p^e-1) is the F-pure threshold

@@

If neither endpoint is the $F$-pure threshold and {\tt Attempts} is set to more than {\tt 2}, then  additional checks are performed at certain numbers within the interval.
A number in the interval is selected, according to criteria specified by the option \textcolor{red}{{\tt GuessStrategy}}, and the function {\tt compareFPT} is used to test that number. If this ``guess'' is correct, its value is returned. Otherwise, the information returned by {\tt  compareFPT} is used to narrow down the interval, and this process is repeated as many times as specified by the value of {\tt Attempts}.

@@

i20 : fpt(f, DepthOfSearch => 3, Attempts => 4) -- more attempts *always*
      improve the estimate

        1   4
o20 = {--, --}
       20  75

@@

i21 : fpt(f, DepthOfSearch => 3, Attempts => 4, GuessStrategy => denominator) 
      -- the exact answer, after a change of strategy

@@

The option {\tt Bounds} allows the user to specify known lower and upper bounds for the $F$-pure threshold of $f$, in order to speed up computations or to refine previously obtained estimates.

@@

i67 : f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;

i68 : fpt(f, DepthOfSearch => 3, Attempts => 5)

        3   1
o68 = {--, --}
       40  13

o68 : List

i69 : fpt(f, DepthOfSearch => 3, Attempts => 5, Bounds => oo)

       48
o69 = ---
      625

o69 : QQ
\end{MyVerbatim}
}
\medspace

% f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;
% fpt(f, DepthOfSearch => 3, Attempts => 5)
% fpt(f, DepthOfSearch => 3, Attempts => 5, Bounds => oo)

% If {\tt guessFPT} is unsuccessful and {\tt UseFSignature} is set to {\tt true}, then the {\tt fpt} function proceeds to use the convexity of the $F$-signature function and a secant line argument to attempt to narrow down the interval bounding the $F$-pure threshold.
% 
% {\small
% \setstretch{.67}
% \begin{MyVerbatim}
% 
% i22 : f = x^5*y^6*(x + y)^9*(x^2 + y^3)^4;
% 
% i23 : fpt(f, DepthOfSearch => 3)
% 
%         2   1
% o23 = {--, --}
%        31  14
% 
% o23 : List
% 
% i24 : fpt(f, DepthOfSearch => 3, UseFSignature => true)
% 
%         181   1
% o24 = {----, --}
%        2750  14
% 
% o24 : List
% 
% i25 : numeric o23
% 
% o25 = {.0645161, .0714286}
% 
% o25 : List
% 
% i26 : numeric o24 -- UseFSignature sharpened the estimate a bit
% 
% o26 = {.0658182, .0714286}
% 
% o26 : List
% \end{MyVerbatim}
% }
% \medspace
% 
% When {\tt FRegularityCheck} is set to {\tt true} and no exact answer has been found, a final check is run to verify whether the final lower bound for the $F$-pure threshold is the exact answer (if it has not already been eliminated as a possibility).
% 
% {\small
% \setstretch{.67}
% \begin{MyVerbatim}
% 
% i27 : f = (x + y)^4*(x^2 + y^3)^6;
% 
% i28 : fpt( f, Attempts => 2, DepthOfSearch => 3 )
% 
%         3   13
% o28 = {--, ---}
%        31  125
% 
% o28 : List
% 
% i29 : fpt(f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true)
% 
%       -- UseFSignature improves the answer a bit
% 
%         1   13
% o29 = {--, ---}
%        10  125
% 
% o29 : List
% 
% i30 : fpt(f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true,
% 
%       FRegularityCheck => true) -- FRegularityCheck finds the answer
% 
%        1
% o30 = --
%       10
% 
% o30 : QQ
% \end{MyVerbatim}
% }
% \medspace
% 
% The computations performed when {\tt UseFSignature} and {\tt FRegularityCheck} are set to {\tt true} are often slow, and often fail to improve the estimate, and for this reason, these options should be used sparingly.
%             It is often more effective to increase the values of {\tt Attempts} or {\tt DepthOfSearch} instead.
% 
% {\small
% \setstretch{.67}
% \begin{MyVerbatim}
% 
% i31 : f = x^7*y^5*(x + y)^5*(x^2 + y^3)^4;
% 
% i32 : timing numeric fpt(f, DepthOfSearch => 3, UseFSignature => true,
% 
%       FRegularityCheck => true)
% 
% o32 = {.0733061, .0769231}
%       -- 2.68899 seconds
% 
% o32 : Time
% 
% i33 : timing numeric fpt(f, Attempts => 5, DepthOfSearch => 3) -- a better
% 
%       answer in less time
% 
% o33 = {.075, .0769231}
%       -- .893389 seconds
% 
% o33 : Time
% 
% i34 : timing fpt(f, DepthOfSearch => 4) -- the exact answer in even less
%       time
% 
%        48
% o34 = ---
%       625
%       -- .361882 seconds
% 
% o34 : Time
% \end{MyVerbatim}
% }
% \medspace

If {\tt guessFPT} is unsuccessful and {\tt FinalAttempt} is set to {\tt true}, the fpt function proceeds to use the convexity of the $F$-signature function and a secant line argument to attempt to narrow down the interval bounding the $F$-pure threshold.
If successful, the new lower bound may coincide with the upper bound, in which case we can conclude that it is the desired $F$-pure threshold.
If that is not the case, an $F$-regularity check is done at the new lower bound, to verify if it is the $F$-pure threshold.

@@

i2 : f = (y - (x-1)^2)*y^2; -- SNC at the origin, but not globally

@@

i4 : fpt(f)

     1
o4 = -
     2

i5 : isSimpleNormalCrossing(f, IsLocal=>false)

o5 = false

i6 : g = (y-1)^2 + (x-7)^2 -- doesn't even pass through the origin

       2    2
o6 = x  + y  - 14x - 2y + 50

o6 : R

i7 : isSimpleNormalCrossing(g)

o7 = true

i8 : isSimpleNormalCrossing(g, IsLocal=>false)

o8 = false

i9 : h = x*y^2*(x+1)^3*(y-1)^4; --SNC everywhere

i10 : isSimpleNormalCrossing(h)

o10 = true

i11 : isSimpleNormalCrossing(h, IsLocal=>false)
